---
title: "4: Physical Properties of Rivers"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Compute recurrence intervals for stream discharge 
2. Analyze the effects of watershed disturbance on recurrence intervals and interpret results against the concept of stationarity
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
How is climate change impacting discharge in streams and rivers? What are the new and ongoing challenges faced by these impacts in watersheds?

- changes in types of precipitation
- temperature impacting evapotranspiration
- less snow pack - more drought in summer/less spring flow

## Session Set Up

```{r, message = FALSE}
getwd()

#install.packages("lfstat")

library(tidyverse)
library(dataRetrieval)
library(lubridate)
library(lfstat)


theme_set(theme_classic())
```

## Recurrence Intervals and Exceededence Probability

A **recurrence interval** is the past recurrence of an event, in this case a peak annual discharge measurement of a given magnitude. The value of a recurrence interval corresponds to the average number of years between discharge of a given magnitude. Typically the minimum amount of years required to construct a recurrence interval is 10, but 30 is more robust. A recurrence interval, T, is calculated as: 

$T = (n+1)/m$

where n is the number of years and m is the ranking of an event within the observed period. We add one to n because we are computing the recurrence interval for a discharge event of a given magnitude *or greater*. 

Similarly, we can calculate an **exceedence probability**, or the probability of encountering a discharge event of a given magnitude or greater in any given year: 

$P = 1/T$

This is where the terms "100-year flood" and similar are derived. Remember this is a probability based on past occurrence, not an accurate forecast of how often we will see that event happening. When current patterns of discharge differ from past patterns, we observe **nonstationary** behavior. Nonstationarity results in events that occur more or less frequency than predicted based on the exceedence probability.

### Has Eno River dicharge displayed stationary behavior over the period of record?

Let's import discharge data for the Eno River near Durham for all available dates.

```{r}
EnoDischarge <- readNWISdv(siteNumbers = "02085070",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "", 
                     endDate = "2021-09-30")

names(EnoDischarge)[4:5] <- c("Discharge", "Approval.Code")

attr(EnoDischarge, "variableInfo")
attr(EnoDischarge, "siteInfo")
  
# Build a ggplot
ggplot(EnoDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Year", y = "Discharge (cfs)")
```

We can then compute recurrence intervals based on the first 30 years of data.

```{r}  
EnoDischarge <- EnoDischarge %>%
  mutate(Year = year(Date), 
         WaterYear = water_year(Date, origin = "usgs")) %>%
  filter(WaterYear != "1963")

# Water Year is a factor. We want to re-classify as numeric.
EnoDischarge$WaterYear <- as.numeric(as.character(EnoDischarge$WaterYear))

EnoRecurrence <- 
  EnoDischarge %>%
  filter(WaterYear < 1994) %>%
  group_by(WaterYear) %>%
  summarise(PeakDischarge = max(Discharge)) %>% 
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval = (length(WaterYear) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)

ggplot(EnoRecurrence, aes(x = WaterYear, y = PeakDischarge)) +
  geom_bar(stat = "identity") +
  labs(x = "Year", y = "Peak Discharge (cfs)")

```

Let's display and model the relationship between peak annual disharge and recurrence interval. We can use the statistical model to compute discharge for recurrence intervals that occur above the 30-year mark.
```{r}
ggplot(EnoRecurrence, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() +
  scale_x_log10() +
  geom_smooth(method = "lm", color = "black", se = FALSE)

Eno.RImodel <- lm(data = EnoRecurrence, PeakDischarge ~ log10(RecurrenceInterval))
summary(Eno.RImodel) # R2 = 91%, pretty good; less reliable further out

#What is the discharge for a 100-year flood in this system? a 500-year flood?
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log10(100) # 9,826
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log10(500) # 12,773
```

What if we were to build a recurrence interval model for the most recent 30 years? How would this compare to the early period recurrence interval?
```{r}
EnoRecurrence.Late <- 
  EnoDischarge %>%
  filter(WaterYear >= 1992) %>%
  group_by(WaterYear) %>%
  summarise(PeakDischarge = max(Discharge)) %>% 
  mutate(Rank = rank(-PeakDischarge), 
         RecurrenceInterval = (length(WaterYear) + 1)/Rank, 
         Probability = 1/RecurrenceInterval)

ggplot(EnoRecurrence, aes(x = RecurrenceInterval, y = PeakDischarge)) +
  geom_point() +
  geom_point(data = EnoRecurrence.Late, color = "#02818a",
             aes(x = RecurrenceInterval, y = PeakDischarge)) +
  scale_x_log10()

Eno.RImodel.Late <- lm(data = EnoRecurrence.Late, PeakDischarge ~ log10(RecurrenceInterval))
summary(Eno.RImodel.Late)
Eno.RImodel.Late$coefficients
Eno.RImodel$coefficients

Eno.RImodel.Late$coefficients[1] + Eno.RImodel.Late$coefficients[2]*log10(100) # 13,762
Eno.RImodel.Late$coefficients[1] + Eno.RImodel.Late$coefficients[2]*log10(500) # 18,176

Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log10(100)
Eno.RImodel$coefficients[1] + Eno.RImodel$coefficients[2]*log10(500)

```

What differences did you see for the recurrence intervals built under different periods of record? How would your prediction of flood events differ if you were to use these models for forecasting purposes? 

> For the later period, the peak discharge at longer recurrence intervals is generally higher. You would expect flood events to be more extreme (or an equivalent amount of discharge to occur with greater frequency).

What would you recommend for a watershed manager seeking to build the most accurate recurrence interval model for the Eno River?

> Either look at the more recent interval only, or look at the full time period, but  weight the more recent period or make more conservative estimates because the findings suggest that precipitation events are becoming more extreme.


### Examining the effects of urbanization on discharge

Salado Creek is located in San Antonio, Texas, an area that has been rapidly urbanizing over the course of the last several decades (http://worldpopulationreview.com/us-cities/san-antonio-population/#byPopulation). Is this system exhibiting stationarity?

Very steep peaks - event based. Appears to have more extreme events in the last 30 years, which makes sense in an urbanizing area. 

```{r}
# Import data
SaladoDischarge <- readNWISdv(siteNumbers = "08178700",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "")
names(SaladoDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(SaladoDischarge, "siteInfo")
  
ggplot(SaladoDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Year", y = "Discharge (cfs)")


```

### Examining the effects of dam construction on recurrence intervals

The stream gage in the Green River near Auburn, Washington, is located directly downstream of the Howard A. Hanson Dam. The dam was built in 1961 for flood control purposes, and the reservoir now provides water supply to the city of Tacoma. How have peak discharges changed since the construction of the dam?

Peak discharges have become more steady/less extreme since the dam construction. This makes sense given that the dam was built for flood control.

```{r}
GreenDischarge <- readNWISdv(siteNumbers = "12113000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "")
names(GreenDischarge)[4:5] <- c("Discharge", "Approval.Code")
attr(GreenDischarge, "siteInfo")
  
ggplot(GreenDischarge, aes(x = Date, y = Discharge)) +
  geom_line() +
  labs(x = "Year", y = "Discharge (cfs)")

```

## Bonus content: Flow Duration Curves and Low Flow Statistics

Flow-duration curves can be generated from daily discharge data, similar to how we calculated recurrence intervals for annual data. 

$P = 100*(m/(n+1))$

where P is the exceedance probability, m is the ranking of all daily mean flows in the period of record (at least 10 years), and n is the total number of daily mean flows. 

We focused today on recurrence intervals, which use peak flow statistics. On the other end of the discharge gradient are low flow statistics, most commonly estimated by 7Q2 and 7Q10 metrics (7-day, 2-year and 10-year annual low flow statistics). These can be used to evaluate drought conditions and are another metric for evaluating stationarity in rivers and streams. 

See the USGS description of these statistics here:(Calculating Flow-Duration and Low-Flow Frequency Statistics at Streamflow-Gaging Stations)[https://pubs.usgs.gov/sir/2008/5126/section3.html]


```{r}
# Best way to make a legend
ggplot(cars, aes(x = speed)) +
  geom_line(aes(y = dist, color = "total discharge")) + 
  scale_color_manual(values = c("total discharge" = "gray"))
#total discharge is just a name here to reference the color

#or to label by a factor column:
ggplot(cars, aes(x = speed)) +
  geom_line(aes(y = dist, color = lake)) + 
  scale_color_manual(values = c("gray","green"))
#total discharge is just a name here to reference the color
```
